Naturalness and Long-Lived Beyond the Standard Model Particles

### 2017-08-02 tex/draft.tex
Add most references

    All SM particles transform according to the gauged symmetry group\fxnote{be good at describing the meaning of `gauged symmetry group'} $\SUthreecSUtwoLUoneY$, where $\SUthreec$ is associated with color charge, $\SUtwoL$ is associated with weak isospin, and $\UoneY$ is associated with weak hypercharge. The gauge bosons are associated with the generators of these groups, thus there are eight gluons associated with $\SUthreec$, three W bosons associated with $\SUtwoL$, and one B boson associated with $\UoneY$. The physical gauge bosons, which are linear combinations of the Ws and B, are the $\Wp$, $\Wm$, $\Z$, and photon ($\gamma$). The fermions are catagorized by which representation of each group they fit into\fxnote{make sure I'm good at explaining this}. For example, $\SUthreec$ distinguishes the quarks (which live in the fundamental representation) from the leptons (which are singlets), and $\SUtwoL$ distinguishes the left-handed fermions (doublets) from the right-handed fermions (singlets). Finally, weak hypercharge further distignuishes SM particles according to the Gell-Mann--Nishijima formula, $Y=Q-I_3$, where Q is electric charge and $I_3$ is the third component of weak isospin.

    Experimentally, the $\Wpm$, $\Z$, quarks, and charged leptons are massive\fxnote{Gloss over neutrino mass?}. Explicit mass terms for the fermions or gauge bosons would violate the chiral and gauge symmetries of the SM Lagrangian, however, so $\SUtwoLUoneY$ must be broken. Specifically, the nonzero vacuum expectation value (VEV) of the Higgs field spontaneously breaks $\SUtwoLUoneY$ to $\UoneQ$. In the process, three of the four degrees of freedom associated with the complex Higgs doublet become the longitudinal polarizations of $\Wpm$ and $\Z$, giving them mass. The remaining degree of freedom is the physical Higgs boson. Expanding the Higgs field about its VEV produces Yukawa interactions such as those in Eq.~\ref{yukawa_terms}. These terms are interpreted as giving mass to the quarks and charged leptons mass, where the mass of each particle is given by the product of the Higgs VEV, $v$, and the coupling constant, $y$.

@@

    SM parameters, such as particle masses\fxnote{what else is corrected. How?}, are corrected by interactions with virtual particles through loop diagrams. Fig.~\ref{loop_diagrams}, for example, shows corrections to the Higgs and electron masses. Because the Higgs is a fundamental scalar, its corrections depend quadratically on the cutoff scale, $\Lambda$, which is the largest energy scale for which the SM is valid. The fermions and gauge bosons, on the other hand, only depend logarithmically\fxnote{check. be able to work out example} on $\Lambda$. The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the bare Higgs mass and the terms that correct it.

    \noindent \begin{figure}[htbp] \begin{center}
        \feynmandiagram [layered layout, horizontal=b to c] {
            a -- [scalar] b,
            b -- [fermion, half left] c,
            c -- [fermion, half left] b,
            c -- [scalar] d,
        };
        \qquad
        \feynmandiagram [layered layout, horizontal=b to c] {
            a -- [fermion] b,
            b -- [fermion] c,
            b -- [photon, half left] c,
            c -- [opacity=0.0, half left] b,
            c -- [fermion] d,
        };
        \caption{Higgs mass (left) corrected by fermion loop and fermion mass (right) corrected by photon loop.\fxnote{cite tikz-feyman}}
        \label{loop_diagrams}
    \end{center} \end{figure}

\section{Naturalness and the Higgs mass}
    \fxnote{cite guidice naturally and t'hooft naturalness}
    \fxnote{Never mentioned dimensional analysis}

    The naturalness criterion can be formulated in many ways, a few of which are reproduced here: (1) All dimensionless parameters should be of order one, (2) An effective theory should not be sensitive to small variations in the parameters of the higher-energy theory, and (3) any dimensionless parameter much smaller than unity must be protected by a custodial symmetry. Physicists implicitly, and perhaps unwittingly, apply naturalness in the form of (1) any time they use dimensional analysis to estimate a quantity, while (2) merely restates\fxnote{verb-subject agreement} the idea that a predictive effective theory requires low-energy phenomena to be decoupled from the details of the high-energy theory\fxnote{This is my weakest point because I don't understand it as well as I'd like to. It's also wordy and maybe repetative}. This discussion focuses on statement (3), which is in some ways a more nuanced version of (1) that explains cases where dimensional analysis fails.

    All massive particles correct the Higgs mass with terms proportional to $\Lambda^2$. Experimentally, the Higgs mass is \SI{125}{\giga\electronvolt}\fxnote{cite CMS, ATLAS}, a value that must be achieved by the sum of the bare mass and all corrections. Although corrections differ in sign, the precise cancellation of many extremely large terms to produce a value as small as the Higgs mass is extremely unlikely. In fact, the fine-tuning necessary to produce the Higgs mass is on the order of one part in \num{e30} if the SM is valid up to the Plank scale ($\mPlanck\approx$ \SI{e16}{\giga\electronvolt}), where quantum gravity becomes important. While it's possible that the bare mass and corrections just so happen to almost perfectly cancel, such miraculous fine tuning could also be a hint that some deeper physical mechanism is at work. That is, either some previously unknown symmetry is protecting the Higgs mass or the cutoff scale is order(TeV).

    As a fundamental scalar, the Higgs is doubly cursed. As explained in section \ref{SM}, $\mHiggs$ is quadratically sensitive to $\Lambda$, but that's not all: the Higgs field also lacks the chiral and gauge symmetries enjoyed by the fermions and gauge bosons. These symmetries act as the custodial symmetries mentioned in naturalness statement (3) in that they protect the fermion and gauge boson masses from large corrections. Because the SM chiral and gauge symmetries only apply to the massless theory, the SM symmetry group would be enlarged if fermions and gauge bosons were massless. That is, if the fermions and gauge bosons were massless at tree level, the chiral and gauge symmetries would cause their masses to be uncorrected to all orders in perturbation theory. As massive particles, however, the chiral and gauge symmetries instead guarantee that any corrections to their tree-level masses are proportional to their tree-level masses, preventing any large fine tuning between correction terms.\fxnote{wordy as hell}

    \fxnote{Clarify that not technically a problem. Nature might just be that way. Anthropic principle almost like an attempt to restore naturalness to a universe that appears unnatural, says Andrew} There are three main approaches to the apparently unnatural Higgs mass\fxnote{specify}: (1) argue that it was never really a problem in the first place, (2) reduce the SM cutoff scale, and (3) introduce a new symmetry to protect the Higgs mass. Arguments against naturalness often involve the anthropic principle, which posits that it should come as no surprise that humans live in a universe whose fundamental parameters are tuned to produce carbon-based life. Alternatively, some theories, such as large extra dimensions, lower the SM cutoff by proposing that gravity is spread across more than three spatial dimensions, which lowers the Planck mass itself. Finally, the dominant\fxnote{what do I mean by dominant?} BSM theory that aims to solve the naturalness problem by introducing a new symmetry is supersymmetry, which is explained below.

\section{Supersymmetry}
    \fxnote{seems like I should explicitly mention that naturalness favors lighter sparticles}  
    Supersymmetry (SUSY) introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between bosons and fermions\fxnote{I need to be better at explaining exactly what this means}. In its simplest form, SUSY results in one new boson for every SM fermion, one new fermion for every SM boson, and one new Higgs doublet.\fxnote{make sure I understand the new Higgs doublet} The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-$\frac{1}{2}$ superpartners to the SM bosons are given the name of their SM counterpart with ``ino" tacked on to the end (e.g. Higgsino or wino).\fxnote{describe charginos, neutralinos?} Collectively the superpartners to the SM particles are known as sparticles.\fxnote{cite martin}

    When calculating contributions from loop diagrams, one finds that fermion loops differ in sign from boson loops, which means that in SUSY every bosonic correction to the Higgs mass is cancelled by a fermionic correction and vice versa. In fact, exact supersymmetry predicts that the bare Higgs mass is uncorrected to all orders in perturbation theory, as shown in Fig.~\ref{exact_susy_correction} below.

    \fxnote{Fix diagram} 
    \noindent \begin{figure}[htbp] \begin{center}
        \feynmandiagram [inline=(d.base), layered layout, medium, horizontal=b to c] {
        a -- [scalar] b,
        b -- [fermion, half left] c,
        c -- [fermion, half left] b,
        c -- [scalar] d,
    };
    \quad + \quad
        \begin{tikzpicture}
            \begin{feynman}
                \diagram [baseline={(current bounding box.center)}, horizontal=a to c] {
                    a -- [scalar] b,
                    b -- [scalar] c,
            };
                \draw (b) arc [charged scalar, start angle=270, end angle=90, radius=0.7cm];
            \end{feynman}
        \end{tikzpicture}
    \quad = \quad 0 \quad
    \caption{Broken!\fxnote{cite something}}
    \label{exact_susy_correction}
    \end{center} \end{figure}

    Exact SUSY also requires that sparticles have the same mass as their SM counterparts, so the lack of sparticles in collider searches implies that SUSY must be broken. With broken susy, the same diagram as above produces a correction of the form\fxnote{cite feng}

@@

\section{Long-Lived Particles}
    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and \fxnote{and/or?} produce a signal that could be missed by common SUSY search techniques. Specifically, we are interested in SUSY models that predict new long-lived particles (LLPs), which are defined as particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point. This category includes everything from particles that decay at \fxnote{give ballparky number ($\approx$100um?)} to particles that propagate through the entire detector.

    In the SM and in BSM theories, LLPs result from any of the following conditions: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly-virtual intermediate states, or (4) limited phase space in which to decay. In the SM, charge and baryon number conservation guarantee the absolute stability of the electron and proton as the the lightest particles carrying their respective conserved quantities. Neutrons and muons, whose decays are shown in Fig.\ref{sm_llp_processes}, are long-lived for reasons (2), (3), and (4). Both are pure weak decays, which gives them a smaller coupling constant, but they also decay through a W-boson, which is far more massive than either parent particle, and most importantly, the proton and electron masses leave them little phase space in which to decay. \fxnote{Mention neutrino, pion, kaon, some other bound state I might be forgetting?}\fxnote{Add diagrams?}

@@

                \diagram* {
                    (n1a) -- [fermion] (n1b) -- [fermion] (p1),
                    (n2) -- [fermion] (p2),
                    (n3) -- [fermion] (p3),
                    (n1b) -- [boson, edge label=\(\W^{-}\)] (w),
                    (w) -- [fermion] (lc),
                    (ln) -- [fermion] (w),
                };
                \draw [decoration={brace}, decorate] (n3.south west) -- (n1a.north west)
                node [pos=0.5, left] {\(n\)};
                \draw [decoration={brace}, decorate] (p1.north east) -- (p3.south east)
                node [pos=0.5, right] {\(p\)};
            \end{feynman}
            \end{tikzpicture}
            \qquad\qquad
            \feynmandiagram [layered layout, horizontal=a to b] {
                a [particle=\(\mu\)] -- [fermion] b,
                b -- [fermion] d [particle=\(\nu_{\mu}\)],
                b -- [boson, edge label=\(\W^{-}\)] c,
                c -- [fermion] e [particle=\(e\)],
                c -- [anti fermion] f [particle=\(\nu_{e}\)],
            };
            \caption{Neutron and muon decays.}
            \label{sm_llp_process}
        \end{center} \end{figure}

    In SUSY models, LLPs arise for the same basic reasons, and different SUSY models produce LLPs in different ways. The most straightforward SUSY LLP occurs in models with conserved R-partity, a proposed symmetry that prevents proton decay\fxnote{Be prepared to show this. Maybe add some diagrams illustrating proton decay w/o R-parity}. In models with R-parity, the SM particles are assigned R-parity +1 and their superpartners are assigned R-parity -1. Conserving the product of R-parity at each vertex has two immediate phenomenalogical consequences: (1) sparticles can only be produced in pairs and (2) the lightest supersymmetric particle (LSP) is absolutely stable. A neutral LSP could show up in collider searches as missing energy ($\MET$), making $\MET$ a key signature in many conventional SUSY searches.

    On the other hand, models with weakly coupled R-parity violating (RPV) terms produce long-lived but not perfectly stable LSPs. A similar situation arises in GMSB models\fxnote{should I describe GMSB?} where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which supresses the NSLP decay rate and can lead to LLPs.\fxnote{cite dimopoulos low energy}

@@

    LLPs can also arise from particular SUSY mass spectra. Models in the Split SUSY paradigm, for example, propose that the scalar sparticles are significantly more massive than the gauginos. In these models, the gluino becomes long-lived when its decay to a quark and neutralino is mediated by a highly-virtual squark. \fxnote{cite kilian} \fxnote{mention that Split throws out naturalness motivation?} On the other hand, some AMSB models\fxnote{explain AMSB?} predict the winos as the lightest gauginos, which leads to nearly-degenerate LSP and NLSP. The NLSP wino then becomes long-lived when its decay is suppressed by the lack of available phase space. \fxnote{cite R\&S out of this world via disappearing}

@@

\section{The CMS Detector}\fxnote{Add pictures of everything}
    \subsection{Overview}
        \fxnote{define eta, coordinate system in general}
        \fxnote{Mention that ATLAS is similar, says Andrew}  
        \fxnote{Maybe mention PDFs. Definitely know about them, says Andrew}  
        \fxnote{cite cms experiment at the cern lhc}
        The Compact Muon Solenoid (CMS) detector is designed to reconstruct a variety of particles from collisions at the Large Hadron Collider (LHC), which is a circular proton-proton collider that ran with center-of-mass energy \SI{7}{\tera\electronvolt} in 2010-11, \SI{8}{\tera\electronvolt} in 2012, and \SI{13}{\tera\electronvolt} from 2015 to the present. To meet this goal, CMS combines information from several subdetectors nested radially in and around a \SI{4}{\tesla} superconducting solenoid. All together, CMS is \SI{21.6}{\m} long, \SI{14.6}{\m} across, and weighs approximately \SI{12500}{t}. The subdetectors that make up CMS are shown in Fig.~\ref{cms} and described below.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.7\textwidth]{images/cms}
            \caption{CMS}
        \label{cms}
        \end{center} \end{figure}

    \subsection{Tracker}
        \fxnote{cite CMS TDR, jinst 12 c02033}
        \fxnote{mention that analyses below used pre-upgrade tracker}
        In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions. Inside \SI{20}{cm}, the pixel tracker  uses smaller\fxnote{than what?} silicon pixels to keep the occupancy near \SI{0.1}{\percent} despite the high\fxnote{specify} particle flux. Outside \SI{20}{\cm}, the silicon strip tracker (SST) achieves similar occupancy with larger silicon strips.\fxnote{check occupacy numbers}

        The pixel tracker, which was upgraded in winter 2016/17, makes up the innermost layer and features approximately \num{124} million silicon pixels, each of which covers an active area of \num{100} by \SI{150}{\micro\meter^2}. The pixels are spread between four barrel layers (BPix) at $\num{2.9}<\mathrm{r}<\SI{16}{cm}$ and 3 endcap disks (FPix) at $\num{29} < \mathrm{z} < \SI{52}{cm}$. The upgraded pixel tracker, shown in Fig.~\ref{pixel_old_new}, should provide 4-hit tracking out to $\eta < \num{2.4}$ and improved b-tagging over the original pixel tracker despite the increased pileup expected in 2017.

        \noindent \begin{figure}[htbp] \begin{center}
            \includegraphics[width=0.5\textwidth]{images/pixel_old_new}
            \caption{Diagram showing barrel pixel layers before and after the winter 2016/17 upgrade.\fxnote{cite website found through Andrew's citations}}
            \label{cms}
        \end{center} \end{figure}

        Just outside the pixel tracker, the silicon strip tracker (SST) occupies the $\num{20} < r < \SI{116}{cm}$ and $\lvert z \rvert < \SI{282}{cm}$ region of CMS. Composed of \num{9.3} million silicon strips that vary from \_ to \_ in size \fxnote{look up} depending on particle flux in the region, the SST covers an active area of roughly \SI{196}{m^2}. At each trigger, the strip output voltage is read out via optical link.

    \subsection{Electromagnetic Calorimeter}
        \fxnote{cite CMS TDR}
        \fxnote{I define EE \& EB but never use those definitions}
        After traversing the inner tracker, particles next encounter the electromagnetic calorimeter (ECAL). As a homogeneous scintillation calorimeter, ECAL uses \num{61200} lead tunstate crystals in the barrel and \num{7324} in each endcap to reconstruct the energy deposited during electromagnetic showers. Lead tungstate crystals allow for a fast (\SI{80}{\percent} of light within \SI{25}{ns}), compact (radiation length = \SI{0.89}{cm}), fine-grained (Moli\`ere radius = \SI{2.2}{cm}), and radiation hard (up to 10 Mrad\fxnote{learn context for this number}) calorimeter. The main drawback is the relatively low light yield (\SI{30}{photon\per\mega\electronvolt}), which necesitates photodetectors with intrinsic gain that work in magnetic fields.

>>>>

    All SM particles transform according to the gauged symmetry group\fxnote{be good at describing the meaning of `gauged symmetry group'} $\SUthreecSUtwoLUoneY$, where $\SUthreec$ is associated with color charge, $\SUtwoL$ is associated with weak isospin, and $\UoneY$ is associated with weak hypercharge. The gauge bosons are associated with the generators of these groups, thus there are eight gluons associated with $\SUthreec$, three W bosons associated with $\SUtwoL$, and one B boson associated with $\UoneY$. The physical gauge bosons, which are linear combinations of the Ws and B, are the $\Wp$, $\Wm$, $\Z$, and photon ($\gamma$). The fermions are catagorized by which representation of each group they fit into\fxnote{make sure I'm good at explaining this}. For example, $\SUthreec$ distinguishes the quarks (which live in the fundamental representation) from the leptons (which are singlets), and $\SUtwoL$ distinguishes the left-handed fermions (doublets) from the right-handed fermions (singlets). Finally, weak hypercharge further distignuishes SM particles according to the Gell-Mann--Nishijima formula, $Y=Q-I_3$, where Q is electric charge and $I_3$ is the third component of weak isospin~\cite{sm_intro}.

    Experimentally, the $\Wpm$, $\Z$, quarks, and charged leptons are massive\fxnote{Gloss over neutrino mass?}. Explicit mass terms for the fermions or gauge bosons would violate the chiral and gauge symmetries of the SM Lagrangian, however, so $\SUtwoLUoneY$ must be broken. Specifically, the nonzero vacuum expectation value (VEV)\fxnote{maybe just use $v$} of the Higgs field spontaneously breaks $\SUtwoLUoneY$ to $\UoneQ$. In the process, three of the four degrees of freedom associated with the complex Higgs doublet become the longitudinal polarizations of $\Wpm$ and $\Z$, giving them mass. The remaining degree of freedom is the physical Higgs boson. Expanding the Higgs field about its VEV produces Yukawa interactions such as those in Eq.~\ref{yukawa_terms}. These terms are interpreted as giving mass to the quarks and charged leptons mass, where the mass of each particle is given by the product of the Higgs VEV, $v$, and the coupling constant, $y$~\cite{sm_intro}.

@@

    SM parameters, such as masses and couplings, are corrected by interactions with virtual particles through loop diagrams. Fig.~\ref{loop_diagrams}, for example, shows interactions that correct the Higgs and electron masses. Because the Higgs is a fundamental scalar, its corrections depend quadratically on the cutoff scale, $\Lambda$, which is the largest energy scale for which the SM is valid. The fermions and gauge bosons, on the other hand, only depend logarithmically\fxnote{check. be able to work out example} on $\Lambda$~\cite{dine_naturalness}. The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the bare Higgs mass and the terms that correct it.

    \noindent \begin{figure}[htbp] \begin{center}
        \feynmandiagram [layered layout, horizontal=b to c] {
            a -- [scalar] b,
            b -- [fermion, half left] c,
            c -- [fermion, half left] b,
            c -- [scalar] d,
        };
        \qquad
        \feynmandiagram [layered layout, horizontal=b to c] {
            a -- [fermion] b,
            b -- [fermion] c,
            b -- [photon, half left] c,
            c -- [opacity=0.0, half left] b,
            c -- [fermion] d,
        };
        \caption{Higgs mass (left) corrected by fermion loop and fermion mass (right) corrected by photon loop. Diagram created with~\cite{tikz}}
        \label{loop_diagrams}
    \end{center} \end{figure}

\section{Naturalness and the Higgs mass}
    The naturalness criterion can be formulated in many ways, a few of which are reproduced here: (1) All dimensionless parameters should be of order one, (2) An effective theory should not be sensitive to small variations in the parameters of the higher-energy theory, and (3) any dimensionless parameter much smaller than unity must be protected by a custodial symmetry~\cite{giudice_naturally, thooft_naturalness}. Physicists implicitly, and perhaps unwittingly, apply naturalness in the form of (1) any time they use dimensional analysis to estimate a quantity, while (2) merely restates the idea that a predictive effective theory requires low-energy phenomena to be decoupled from the details of the high-energy theory. \fxnote{wordy and maybe repetative}. This discussion focuses on statement (3), which is in some ways a more nuanced version of (1) that also explains cases where dimensional analysis fails.

    All massive particles correct the Higgs mass with terms proportional to $\Lambda^2$~\cite{dine_naturalness}\fxnote{show diagrams and terms?}. Experimentally, the Higgs mass is \SI{125}{\giga\electronvolt}~\cite{cms_higgs, atlas_higgs}, a value that must be achieved by the sum of the bare mass and all corrections. Although corrections differ in sign, the precise cancellation of many extremely large terms to produce a value as small as the Higgs mass is extremely unlikely. In fact, the fine-tuning necessary to produce the Higgs mass is on the order of one part in \num{e30} if the SM is valid up to the Plank scale ($\mPlanck\approx$ \SI{e16}{\giga\electronvolt}), where quantum gravity becomes important~\cite{giudice_naturally}. While it's possible that the bare mass and corrections just so happen to almost perfectly cancel, such miraculous fine tuning could also be a hint that some deeper physical mechanism is at work. That is, either some previously unknown symmetry is protecting the Higgs mass or the cutoff scale is $\mathcal{O}\left(\SI{}{\tera\electronvolt}\right)$\fxnote{explain better and cite something}.

    As a fundamental scalar, the Higgs is doubly cursed. As explained in section \ref{SM}, $\mHiggs$ is quadratically sensitive to $\Lambda$, but that's not all: the Higgs field also lacks the chiral and gauge symmetries enjoyed by the fermions and gauge bosons. These symmetries act as the custodial symmetries mentioned in naturalness statement (3) in that they protect the fermion and gauge boson masses from large corrections. Because the SM chiral and gauge symmetries only apply to the massless theory, the SM symmetry group would be enlarged if fermions and gauge bosons were massless. That is, if the fermions and gauge bosons were massless at tree level, the chiral and gauge symmetries would cause their masses to be uncorrected to all orders in perturbation theory. As massive particles, however, the chiral and gauge symmetries instead guarantee that any corrections to their tree-level masses are proportional to their tree-level masses, preventing any large fine tuning between correction terms~\cite{giudice_naturally}.\fxnote{wordy as hell}

    \fxnote{Clarify that not technically a problem. Nature might just be that way. Anthropic principle almost like an attempt to restore naturalness to a universe that appears unnatural, says Andrew} There are three main approaches to the apparently unnatural Higgs mass\fxnote{specify}: (1) argue that it was never really a problem in the first place, (2) reduce the SM cutoff scale, and (3) introduce a new symmetry to protect the Higgs mass~\cite{craig}. Arguments against naturalness often involve the anthropic principle, which posits that it should come as no surprise that humans live in a universe whose fundamental parameters are tuned to produce carbon-based life. Alternatively, some theories, such as large extra dimensions, lower the SM cutoff by proposing that gravity is spread across more than three spatial dimensions, which lowers the Planck mass itself. Finally, the dominant\fxnote{what do I mean by dominant?} BSM theory that aims to solve the naturalness problem by introducing a new symmetry is supersymmetry, which is explained below~\cite{dine_naturalness}.

\section{Supersymmetry}
    \fxnote{seems like I should explicitly mention that naturalness favors lighter sparticles and why the stop correction is so important}

    Supersymmetry (SUSY) introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between bosons and fermions\fxnote{I need to be better at explaining exactly what this means}. In its simplest form, SUSY results in one new boson for every SM fermion, one new fermion for every SM boson, and one new Higgs doublet.\fxnote{make sure I understand the new Higgs doublet} The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-$\frac{1}{2}$ superpartners to the SM bosons are given the name of their SM counterpart with ``ino" tacked on to the end (e.g. Higgsino or wino).\fxnote{describe charginos, neutralinos?} Collectively the superpartners to the SM particles are known as sparticles~\cite{primer}.

    When calculating contributions from loop diagrams, one finds that fermion loops differ in sign from boson loops, which means that in SUSY every bosonic correction to the Higgs mass is cancelled by a fermionic correction and vice versa. In fact, exact supersymmetry predicts that the bare Higgs mass is uncorrected to all orders in perturbation theory, as shown in Fig.~\ref{exact_susy_correction} below~\cite{primer}.

    \fxnote{Fix diagram} 
    \fxnote{Add particle labels} 
    \noindent \begin{figure}[htbp] \begin{center}
        \feynmandiagram [inline=(d.base), layered layout, medium, horizontal=b to c] {
        a -- [scalar] b,
        b -- [fermion, half left] c,
        c -- [fermion, half left] b,
        c -- [scalar] d,
    };
    \quad + \quad
        \begin{tikzpicture}
            \begin{feynman}
                \diagram [baseline={(current bounding box.center)}, horizontal=a to c] {
                    a -- [scalar] b,
                    b -- [scalar] c,
            };
                \draw (b) arc [charged scalar, start angle=270, end angle=90, radius=0.7cm];
            \end{feynman}
        \end{tikzpicture}
    \quad = \quad 0 \quad
        \caption{Broken! Corrections to the Higgs mass from the top (left) and stop (right) cancel in exact SUSY. Diagram created with~\cite{tikz}}
    \label{exact_susy_correction}
    \end{center} \end{figure}

    Exact SUSY also requires that sparticles have the same mass as their SM counterparts, so the lack of sparticles in collider searches implies that SUSY must be broken. With broken SUSY, the same diagram as above produces a correction of the form~\cite{feng}

@@

\section{Long-Lived Particles}
    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and \fxnote{and/or?} produce a signal that could be missed by common SUSY search techniques. Specifically, we are interested in SUSY models that predict new long-lived particles (LLPs), which are defined as particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point. This category includes everything from particles that decay at 100um to particles that propagate through the entire detector\fxnote{I made up 100um. Find real number and cite source}.

    In the SM and in BSM theories, LLPs result from any of the following conditions: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly-virtual intermediate states, or (4) limited phase space in which to decay\fxnote{find source to cite}. In the SM, charge and baryon number conservation guarantee the absolute stability of the electron and proton as the the lightest particles carrying their respective conserved quantities. Neutrons and muons, whose decays are shown in Fig.\ref{sm_llp_processes}, are long-lived for reasons (2), (3), and (4). Both are pure weak decays, which gives them a smaller coupling constant, but they also decay through a W-boson, which is far more massive than either parent particle, and most importantly, the proton and electron masses leave them little phase space in which to decay\fxnote{find something to cite}. \fxnote{Mention neutrino, pion, kaon, some other bound state I might be forgetting?}

@@

                \diagram* {
                    (n1a) -- [fermion] (n1b) -- [fermion] (p1),
                    (n2) -- [fermion] (p2),
                    (n3) -- [fermion] (p3),
                    (n1b) -- [boson, edge label=\(\W^{-}\)] (w),
                    (w) -- [fermion] (lc),
                    (ln) -- [fermion] (w),
                };
                \draw [decoration={brace}, decorate] (n3.south west) -- (n1a.north west)
                node [pos=0.5, left] {\(n\)};
                \draw [decoration={brace}, decorate] (p1.north east) -- (p3.south east)
                node [pos=0.5, right] {\(p\)};
            \end{feynman}
            \end{tikzpicture}
            \qquad\qquad
            \feynmandiagram [layered layout, horizontal=a to b] {
                a [particle=\(\mu\)] -- [fermion] b,
                b -- [fermion] d [particle=\(\nu_{\mu}\)],
                b -- [boson, edge label=\(\W^{-}\)] c,
                c -- [fermion] e [particle=\(e\)],
                c -- [anti fermion] f [particle=\(\nu_{e}\)],
            };
            \caption{Long-lived decays of the neutron and muon. Diagram created with~\cite{tikz}}
            \label{sm_llp_processes}
        \end{center} \end{figure}

    In SUSY models, LLPs arise for the same basic reasons, and different SUSY models produce LLPs in different ways. The most straightforward SUSY LLP occurs in models with conserved R-partity, a proposed symmetry that prevents proton decay\fxnote{Be prepared to show this. Maybe add some diagrams illustrating proton decay w/o R-parity}. In models with R-parity, the SM particles are assigned R-parity +1 and their superpartners are assigned R-parity -1. Conserving the product of R-parity at each vertex has two immediate phenomenalogical consequences: (1) sparticles can only be produced in pairs and (2) the lightest supersymmetric particle (LSP) is absolutely stable. A neutral LSP could show up in collider searches as missing energy ($\MET$), making $\MET$ a key signature in many conventional SUSY searches~\cite{primer}.

    On the other hand, models with weakly coupled R-parity violating (RPV) terms produce long-lived but not perfectly stable LSPs\fxnote{find something to cite}. A similar situation arises in GMSB models\fxnote{should I describe GMSB?} where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which supresses the NSLP decay rate and can lead to LLPs~\cite{dimopoulos_low_energy}.

@@

    LLPs can also arise from particular SUSY mass spectra. Models in the Split SUSY paradigm, for example, propose that the scalar sparticles are significantly more massive than the gauginos. In these models, the gluino becomes long-lived when its decay to a quark and neutralino is mediated by a highly-virtual squark~\cite{kilian_split}. \fxnote{mention that Split throws out naturalness motivation?} On the other hand, some AMSB models\fxnote{explain AMSB?} predict the winos as the lightest gauginos, which leads to nearly-degenerate LSP and NLSP. The NLSP wino then becomes long-lived when its decay is suppressed by the lack of available phase space~\cite{randall_sundrum_out_of_this_world}.

@@

\section{The CMS Detector}\fxnote{Add pictures of everything}
    \subsection{Overview}
        \fxnote{define eta, coordinate system in general}
        \fxnote{Mention that ATLAS is similar, says Andrew}  
        \fxnote{Maybe mention PDFs. Definitely know about them, says Andrew}  
        The Compact Muon Solenoid (CMS) detector is designed to reconstruct a variety of particles from collisions at the Large Hadron Collider (LHC), which is a circular proton-proton collider that ran with center-of-mass energy \SI{7}{\tera\electronvolt} in 2010-11, \SI{8}{\tera\electronvolt} in 2012, and \SI{13}{\tera\electronvolt} from 2015 to the present\fxnote{say better}\fxnote{cite something?}. To meet this goal, CMS combines information from several subdetectors nested radially in and around a \SI{4}{\tesla} superconducting solenoid. All together, CMS is \SI{21.6}{\m} long, \SI{14.6}{\m} across, and weighs approximately \SI{12500}{t}. The subdetectors that make up CMS are shown in Fig.~\ref{cms} and described below~\cite{cms_experiment}.

        \noindent \begin{figure}[htbp] \begin{center}
        \includegraphics[width=0.7\textwidth]{images/cms}
            \caption{The CMS detector\fxnote{Cite that website from Andrew's paper}}
        \label{cms}
        \end{center} \end{figure}

    \subsection{Tracker}
        \fxnote{cite CMS TDR, jinst 12 c02033}
        \fxnote{mention that analyses below used pre-upgrade tracker}
        In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions. Inside \SI{20}{cm}, the pixel tracker  uses smaller\fxnote{than what?} silicon pixels to keep the occupancy near \SI{0.1}{\percent} despite the high\fxnote{specify} particle flux. Outside \SI{20}{\cm}, the silicon strip tracker (SST) achieves similar occupancy with larger silicon strips~\cite{cms_tdr}.\fxnote{check occupacy numbers}

        The pixel tracker, which was upgraded in winter 2016/17, makes up the innermost layer and features approximately \num{124} million silicon pixels, each of which covers an active area of \num{100} by \SI{150}{\micro\meter^2}. The pixels are spread between four barrel layers (BPix) at $\num{2.9}<\mathrm{r}<\SI{16}{cm}$ and 3 endcap disks (FPix) at $\num{29} < \mathrm{z} < \SI{52}{cm}$. The upgraded pixel tracker, shown in Fig.~\ref{pixel_old_new}, should provide 4-hit tracking out to $\eta < \num{2.4}$ and improved b-tagging over the original pixel tracker despite the increased pileup expected in 2017~\cite{cms_pixel_upgrade}.

        \noindent \begin{figure}[htbp] \begin{center}
            \includegraphics[width=0.5\textwidth]{images/pixel_old_new}
            \caption{Diagram showing barrel pixel layers before and after the winter 2016/17 upgrade.\fxnote{cite website found through Andrew's citations}}
            \label{pixel_old_new}
        \end{center} \end{figure}

        Just outside the pixel tracker, the silicon strip tracker (SST) occupies the $\num{20} < r < \SI{116}{cm}$ and $\lvert z \rvert < \SI{282}{cm}$ region of CMS. Composed of \num{9.3} million silicon strips that vary from \_ to \_ in size \fxnote{look up} depending on particle flux in the region, the SST covers an active area of roughly \SI{196}{m^2}. At each trigger, the strip output voltage is read out via optical link~\cite{cms_pixel_upgrade}.

    \subsection{Electromagnetic Calorimeter}
        \fxnote{cite CMS TDR}
        \fxnote{I define EE \& EB but never use those definitions}
        After traversing the inner tracker, particles next encounter the electromagnetic calorimeter (ECAL). As a homogeneous scintillation calorimeter, ECAL uses \num{61200} lead tunstate crystals in the barrel and \num{7324} in each endcap to reconstruct the energy deposited during electromagnetic showers. Lead tungstate crystals allow for a fast (\SI{80}{\percent} of light within \SI{25}{ns}), compact (radiation length = \SI{0.89}{cm}), fine-grained (Moli\`ere radius = \SI{2.2}{cm}), and radiation hard (up to 10 Mrad\fxnote{learn context for this number}) calorimeter. The main drawback is the relatively low light yield (\SI{30}{photon\per\mega\electronvolt}), which necesitates photodetectors with intrinsic gain that work in magnetic fields\cite{cms_experiment, cms_tdr}.
